---
title: "Text prediction via N-gram Stupid Back-off models"
author: 
- name: Valerio Gherardi
  email: vgherard@sissa.it
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sbo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sbo` package provides utilities for building and evaluating next-word prediction functions based on [Stupid Back-off](https://www.aclweb.org/anthology/D07-1090.pdf) [N-gram models](https://en.wikipedia.org/wiki/N-gram) in R. In this vignette, I illustrate the main feature of `sbo`, including in particular: 

* the typical workflow for building a text predictor from a given training corpus, and 
* the evaluation of next-word predictions through a test corpus.

## Functions and classes

The `sbo` package pivots around two (S3) object classes: 

* `kgram_freqs`: A collection of $k$-gram frequency tables, with $k$ up to a given order $N$. These are obtained from a training corpus through the function `get_kgram_freqs()`.
* `sbo_preds`: these objects directly store next-word predictions, allowing for memory compression and fast access. They can be built from `kgram_freqs` objects through `build_sbo_preds()`.

In addition, `sbo` features the function `eval_sbo_preds`, which allows to evaluate prediction accuracy on a new test corpus.

## Building a next-word prediction function with `sbo`

```{r setup, warning=FALSE, message=FALSE}
library(sbo)
```

In this and the next section we will employ the `twitter_train` and `twitter_test` example datasets, included in `sbo` for illustrative purpose:

```{r}
train <- twitter_train
test <- twitter_test
```

These are small samples of $7·10^4$ and $10^4$ entries, respectively, from the "Tweets" Swiftkey dataset fully available [here](https://www.kaggle.com/crmercado/tweets-blogs-news-swiftkey-dataset-4million). Each entry consists of a single tweet in English, *e.g.*:

```{r}
head(train, 3)
```

The prototypical workflow for building a text-predictor in `sbo` goes as follows:

*Step 0 (optional)*. Build a dictionary from training set,  keeping the top $V=1000$ most frequent words: 
```{r}
# N.B.: get_word_freqs(train) stores word counts in a sorted named integer.
word_freqs <- get_word_freqs(train)
dict <- names(word_freqs)[1:1000]
head(dict)
```
Alternatively, if available, one may use a predefined dictionary.

*Step 1*. Obtain $k$-gram frequencies from training corpus:
```{r}
(freqs <- get_kgram_freqs(train, N = 3, dict)) # 'N' is the order of n-grams
```

*Step 2*. Build next-word prediction tables:
```{r}
(preds <- build_sbo_preds(freqs, L = 3)) # L = Number of predictions per input k-gram 
```
The `preds` object now stores all the information needed to generate next-word predictions according to Stupid Back-Off. The argument `L` fixes the number of predictions to retain per input k-gram (here we prune to the top 3 predictions).

At this point we can predict next words from our model, by using `predict` (see `?predict.sbo_preds` for help on the relevant `predict` method):

```{r}
predict(preds, "i love") # a character vector
predict(preds, c("Colorless green ideas sleep", "See you")) # a char matrix
```

Last, but not least, we can employ our model for generating some beautiful non-sense:

```{r}
set.seed(840)
babble(preds)
babble(preds)
babble(preds)
```

If we wish to save the frequency tables, or the final prediction tables, and reload them in a future session, we can easily do this through `save`/`load`, *e.g.*

```{r, eval=FALSE}
save(preds)
load("preds.rda")
```

For convenience, the objects created in this section are preloaded in `sbo` as `twitter_dict`, `twitter_freqs` and `twitter_preds`.

### Some details on `kgram_freqs` and `sbo_preds` class objects
As the name suggests, objects of class `kgram_freqs` store the counts for each $k$-gram observed in the training corpus. These are the building block
for *any* $N$-gram based language model, and indeed text prediction can be performed directly from the `freqs` object obtained above:

```{r}
predict(freqs, "i love")
```

The output contains the full language model information, i.e. the probabilities[^1] for each possible word completion.

On the contrary, `sbo_preds` objects directly store next-word predictions for each $k$-gram prefix ($k=1,\,2,\dots,\,N-1$) observed in the training corpus.
The advantage provided by `sbo_preds` objects for simple text prediction is two-fold:

1. Memory compression. For instance:
```{r}
size_in_MB <- function(x) format(utils::object.size(x), units = "MB")
sapply(list(preds = preds, freqs = freqs), size_in_MB)
```

2. Fast query:
```{r}
chrono_predict <- function(x) system.time(predict(x, "i love"), gcFirst = TRUE)
lapply(list(preds = preds, freqs = freqs), chrono_predict)
```

Both objects store, through attributes, information about the training process. This can be conveniently obtained through the corresponding `summary()` methods, e.g.

```{r}
summary(preds)
```


[^1]: More precisely, these are the normalized (to unity) scores resulting from
the Stupid Back-Off smoothing method.

##### Internal structure of `sbo_preds` object
Here are some details on the current (under development) implementation of `sbo_preds` objects. Consider:

```{r}
head(preds[[3]])
```

The first two columns correspond to the word codes [^2] of $2$-gram prefixes observed in the training corpus, and the other columns code the top $L=3$ predictions for these $2$-grams. When a $2$-gram $w_1 w_2$ is given as input for text prediction, it is first looked for in the prefix columns of `preds[[3]]`. If not found, $w_2$ is looked for in the prefix column of `preds[[2]]`. If this also fails, the prediction is performed without any prefix, that is, we simply `predict()` the `L` most frequent words, stored in:

```{r}
preds[[1]]
```

[^2]: Coded with respect to the rank sorted dictionary `dict` (the codes `0`, `length(dict) + 1` and `length(dict) + 2` correspond to the Begin-Of-Sentence, End-Of-Sentence and Unknown-Word tokens, respectively).

### Text preprocessing

Usually text corpora require preprocessing before word and $k$-gram tokenization can take place. The `.preprocess` argument of `get_kgram_freqs()` allows for an user specified preprocessing function. The default is the minimal `sbo::preprocess()`, and the optimized `get_kgram_freqs_fast(erase = x, EOS = y)` is equivalent to `get_kgram_freqs(.preprocess = sbo::preprocess(erase = x, EOS = y))` (but substantially more efficient).

### Sentence tokenization

Tokenization at the sentence level is required to obtain terminal $k$-grams (i.e. $k$-grams containing Begin-Of-Sentence or End-Of-Sentence tokens). In the training process, sentence tokenization takes place after text preprocessing.

End-Of-Sentence (single character) tokens are specified by the `EOS` argument of `get_kgram_freqs()` and `get_kgram_freqs_fast()`; empty sentences are always skipped. Also, if the input vector `text` has `length(text) > 1`, the various elements of `text` belong to different entries. 

The process of sentence tokenization can also be performed directly, through `sbo::tokenize_sentences()`, but this is not required for use with `get_kgram_freqs*()`.

## Evaluating next-word predictions

This Section leverages, for convenience, on:

```{r message=FALSE, warning=FALSE}
library(dplyr) # installed with `sbo`
```

Once we have built our next-word predictor, we may want to directly test its predictions on an independent corpus. For this purpose, `sbo` offers the function `eval_sbo_preds`, which performs the following test:

1. Sample a single $N$-gram from each sentence of test corpus.
1. Predict next words from the $(N-1)$-gram prefix.
1. Return all predictions, together with the true word completions.

As a concrete example, we test the text-predictor trained in the previous section over the Twitter (independent) test set.

```{r}
set.seed(840)
(eval <- eval_sbo_preds(preds, test)) # test <- sbo::twitter_test
```

As it is seen, `eval_sbo_preds` returns a tibble containing the input $(N-1)$-grams, the true completions, the predicted completions and a column indicating whether one of the predictions were correct or not.

We can estimate predictive accuracy as follows (the uncertainty in the estimate is approximated by the binomial formula $\sigma = \sqrt{\frac{p(1-p)}{M}}$, where $M$ is the number of trials):

```{r}
eval %>% summarise(accuracy = sum(correct)/n(), 
                   uncertainty = sqrt(accuracy * (1 - accuracy) / n())
                   )
```

We may want to exclude from the test $N$-grams ending by the End-Of-Sentence token (here represented by `"<EOS>"`):

```{r}
eval %>% # Accuracy for in-sentence predictions
        filter(true != "<EOS>") %>%
        summarise(accuracy = sum(correct) / n(),
                  uncertainty = sqrt(accuracy * (1 - accuracy) / n())
                  )
```

In trying to reduce the size (in physical memory) of your text-predictor, it might be useful to prune the model dictionary. The following command plots an histogram of the distribution of correct predictions in our test.


```{r}
if (require(ggplot2)) {
        eval %>%
                filter(correct, true != "<EOS>") %>%
                select(true) %>%
                transmute(rank = match(true, table = attr(preds, "dict"))) %>%
                ggplot(aes(x = rank)) + geom_histogram(binwidth = 25)
}
```

Apparently, the large majority of correct predictions come from the first ~ 300 words of the dictionary, so that if we prune the dictionary excluding words with rank greater than, *e.g.*, 500 we can reduce the size of our model without seriously affecting its prediction accuracy.
